---
title: "CS7641 Machine Learning First Assignment "
author: "T. Ruzmetov"
date: "September 20, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```



# Introduction

Purpose of this assignment is to investigate and compare the performance of five different machine
learning techniques such as K-Nearest Neighbor, Decision Trees with pruning, Boosted Derision Trees, Support Vector
Machines and Artificial Neural Networks by applying them to two distinct classification problems(distinct datasets).
In order to fulfill the requirement, "Lending Club Data set" from Kaggle and Adult Data set from UCI machine learning
repository are used. For all methods on both data sets cross validation is performed in order to find optimum hyper parameter. Learning curve experiment is done by altering training data size and monitoring performance via prediction accuracy.    



## Lending Club Dataset

Lending Club is the world’s largest online marketplace connecting borrowers and investors. Getting loan has become common practice in developed countries, which makes it interesting to learn how banks determine customer eligibility or credit worthiness given some information. In this problem, we apply unsupervised learning methods to predict if
customer is going to pay or default on their loan based on provided set of features with labels. Provided data set
contain complete loan data for all loans issued through 2007 to 2015, including the current loan status(Current,
Late, Fully Paid, etc.). Instances with "Current" loan status are discarded to make it simple binary classification
problem with two labels(unpaid,paid). After cleaning and feature extraction, the data set contains 15 features and
285373 samples. 90% of entire data is allocated to training set and 10% is separated out for testing. Then only only
20% of the training set is used for model training due to large size resulting in high time consumption.  

```{r, echo=FALSE}
# put targe distributions
```

## Adult Dataset

The adult data set contains census information from 1994.  Our task is to predict whether a person makes more
than $50K/year. After preprocessing is applied, there are 11 features and 45000 remaining instances. The target 
feature "income" is labeled as "high" when income is greater than $50K and "low" when it is less. 

## K-Neirest Naighbour 

Knn algorithm classifies a data points based on its K closest neighbors in distance, where over represented
class within K will get the vote. K value and distance metric are the only parameters to tune for cross validation.
It is slow with large data sets and high dimensional data. Also categorical features don't work well. 5-fold twice
repeated cross validation by changing k value gave $k_{best}=43$ for Lending Club data set and $k_{best}=27$ for Adult Data set. 
This values of k correspond to maximum accuracy as shown in below plots. Then, the best model is used to make prediction on test
set, where 

\begin{figure}
  \center
    \includegraphics[width=6.0cm]{figs/LC_knn_acc_naigh.pdf}
    \includegraphics[width=6.0cm]{figs/AD_knn_acc_naigh.pdf}
  \center
  \caption{Cross validation plots for KNN model}
  \label{fig:knn_val}
\end{figure}


\begin{figure}
  \center
    \includegraphics[width=8.0cm]{figs/LC_knn_learning_curve.png}
    \includegraphics[width=8.0cm]{figs/AD_knn_learning_curve.png}
  \center
  \caption{Cross validation plots for KNN model}
  \label{fig:lear_curve_knn}
\end{figure}


```{r,echo=FALSE}
library(knitr)
LC_con_mat_knn <- read.table("output/LC_confusion_mat_knn.txt", sep = "", header = TRUE)
AD_con_mat_knn <- read.table("output/AD_confusion_mat_knn.txt", sep = "", header = TRUE,
                             col.names = c("high","low"),row.names = c("high","low"))
AD_con_mat_knn$row.names <- NULL
kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "Confusion matrix for Lending
      Club(left) and Adult data(right)")
```


## Decision Tree Algorithm with Prunning

\includegraphics[width=16.0cm]{figs/LC_pruned_tree_diag.png}

\begin{figure}
  \center
    \includegraphics[width=8.0cm]{figs/LC_tree_cp_xerror.png}
    \includegraphics[width=8.0cm]{figs/AD_tree_cp_xerror.png}
  \center
  \caption{Cross validation plots for Tree model}
  \label{fig:cross_val_tree}
\end{figure}



\includegraphics[width=8.0cm]{figs/LC_tree_learning_curve.png}
\includegraphics[width=8.0cm]{figs/AD_tree_learning_curve.png}


```{r,echo=FALSE}
LC_con_mat_tree <- read.table("output/LC_confusion_mat_tree.txt", sep = "", header = TRUE)
AD_con_mat_tree <- read.table("output/AD_confusion_mat_tree.txt", sep = "", header = TRUE)
kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "Decision Trees confusion matrix for Lending
      Club(left) and Adult data(right)")
```


## Gradient Boosting

\includegraphics[width=8.0cm]{figs/LC_boost_acc_iter_shrink.pdf}
\includegraphics[width=8.0cm]{figs/AD_boost_acc_iter_shrink.pdf}



\includegraphics[width=8.0cm]{figs/LC_boosting_learning_curve.png}
\includegraphics[width=8.0cm]{figs/AD_boosting_learning_curve.png}

```{r,echo=FALSE}
LC_con_mat_boost <- read.table("output/LC_confusion_mat_boost.txt", sep = "", header = TRUE)
AD_con_mat_boost <- read.table("output/AD_confusion_mat_boost.txt", sep = "", header = TRUE)

kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "Boosting confusion matrix for Lending
      Club(left) and Adult Data(right)")
```

## Neural Networks

\includegraphics[width=8.0cm]{figs/LC_nnet_ROC_units_weight.pdf}
\includegraphics[width=8.0cm]{figs/AD_nnet_ROC_units_weight.pdf}



\includegraphics[width=8.0cm]{figs/LC_nnet_learning_curve.png}
\includegraphics[width=8.0cm]{figs/AD_nnet_learning_curve.png}


```{r,echo=FALSE}
LC_con_mat_nnet <- read.table("output/LC_confusion_mat_nnet.txt", sep = "", header = TRUE)
AD_con_mat_nnet <- read.table("output/AD_confusion_mat_nnet.txt", sep = "", header = TRUE)

kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "NNet confusion matrix for Lending
      Club(left) and Adult Data(right)")
```

## Support Vector Machines



## Conclusion



