---
title: "CS7641 Machine Learning First Assignment "
author: "T. Ruzmetov"
date: "September 20, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```



# Introduction

Purpose of this assignment is to investigate and compare the performance of five different machine
learning techniques such as K-Neirest Naighbour, Desicion Trees with prunning, Boosted Desision Trees, Support Vector
Machines and Artificial Neural Networks by applying them to two distinct classification problems(distinct datasets).
In order to fullfil the requirement, "Lending Club Dataset" from Kaggle and Adult Dataset from UCI machine leraning
repository are used. For all methods on both datasets cross validation is performed in order to find optimum hyperparameter. Learning curve experiment is done by altering training data size and monitoring performance via prediction accuracy.    



## Lending Club Dataset

Lending Club is the world’s largest online marketplace connecting borrowers and investors. Getting loan has become common practice in developed countries, which makes it interesting to learn how banks determine customer eligibility or credit worthiness given some information. In this problem, we apply unsupervised learning methods to predict if
customer is goint to pay or default on their loan based on provided set of features with labels. Provided dataset
contain complete loan data for all loans issued through 2007 to 2015, including the current loan status(Current,
Late, Fully Paid, etc.). Instances with "Current" loan status are discarded to make it simple binary classification
problem with two labels(unpaid,paid). After cleaning and feature extraction, the dataset contains 15 features and
285373 samples. 90% of entire data is allocated to training set and 10% is separated out for testing. Then only only
20% of the training set is used for model training due to large size resulting in high time consumption.  

```{r, echo=FALSE}
# put targe distributions
```

## Adult Dataset

The adult dataset contains census information from 1994.  Our taske is to predict whether a person makes more
than $50K/year. After preprocessing is applied, there are 11 features and 45000 remaining instances. The target 
feature "income" is labeled as "high" when income is greater than $50K and "low" when it is less. 


## K-Neirest Naighbour 

Knn algorithm classifies a data points based on its K closest naighbours in distance, where overrepresented
class within K will get the vote. K value and distance metric are the only parameters to tune for cross validation.
It is slow with large datasets and high dimentional data. Also categorical features don't work well.  


\includegraphics[width=6.0cm]{figs/LC_knn_acc_naigh.pdf}
\includegraphics[width=6.0cm]{figs/AD_knn_acc_naigh.pdf}


\includegraphics[width=8.0cm]{figs/LC_knn_learning_curve.png}
\includegraphics[width=8.0cm]{figs/AD_knn_learning_curve.png}



```{r,echo=FALSE}
library(knitr)
LC_con_mat_knn <- read.table("output/LC_confusion_mat_knn.txt", sep = "", header = TRUE)
AD_con_mat_knn <- read.table("output/AD_confusion_mat_knn.txt", sep = "", header = TRUE,
                             col.names = c("high","low"),row.names = c("high","low"))
AD_con_mat_knn$row.names <- NULL
kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "Confusion matrix for Lending
      Club(left) and Adult data(right)")
```


## Decision Tree Algorithm with Prunning

\includegraphics[width=16.0cm]{figs/LC_pruned_tree_diag.png}



\includegraphics[width=8.0cm]{figs/LC_tree_cp_xerror.png}
\includegraphics[width=8.0cm]{figs/AD_tree_cp_xerror.png}



\includegraphics[width=8.0cm]{figs/LC_tree_learning_curve.png}
\includegraphics[width=8.0cm]{figs/AD_tree_learning_curve.png}

```{r,echo=FALSE}
LC_con_mat_tree <- read.table("output/LC_confusion_mat_tree.txt", sep = "", header = TRUE)
AD_con_mat_tree <- read.table("output/AD_confusion_mat_tree.txt", sep = "", header = TRUE)
kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "Decision Trees confusion matrix for Lending
      Club(left) and Adult data(right)")
```


## Gradient Boosting

\includegraphics[width=8.0cm]{figs/LC_boost_acc_iter_shrink.pdf}
\includegraphics[width=8.0cm]{figs/AD_boost_acc_iter_shrink.pdf}



\includegraphics[width=8.0cm]{figs/LC_boosting_learning_curve.png}
\includegraphics[width=8.0cm]{figs/AD_boosting_learning_curve.png}

```{r,echo=FALSE}
LC_con_mat_boost <- read.table("output/LC_confusion_mat_boost.txt", sep = "", header = TRUE)
AD_con_mat_boost <- read.table("output/AD_confusion_mat_boost.txt", sep = "", header = TRUE)

kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "Boosting confusion matrix for Lending
      Club(left) and Adult Data(right)")
```

## Neural Networks

\includegraphics[width=8.0cm]{figs/LC_nnet_ROC_units_weight.pdf}
\includegraphics[width=8.0cm]{figs/AD_nnet_ROC_units_weight.pdf}



\includegraphics[width=8.0cm]{figs/LC_nnet_learning_curve.png}
\includegraphics[width=8.0cm]{figs/AD_nnet_learning_curve.png}


```{r,echo=FALSE}
LC_con_mat_nnet <- read.table("output/LC_confusion_mat_nnet.txt", sep = "", header = TRUE)
AD_con_mat_nnet <- read.table("output/AD_confusion_mat_nnet.txt", sep = "", header = TRUE)

kable(list(LC_con_mat_knn, AD_con_mat_knn),caption = "NNet confusion matrix for Lending
      Club(left) and Adult Data(right)")
```

## Support Vector Machines



## Conclusion



